Anxiety: a decision-theoretic perspective

### 2019-02-03 manuscript/manuscript.tex
figure updates for manuscript

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/01_field.png}}%
  }
  \par Figure 1: The open field environment is a large deterministic gridworld sans one rewarding
  (blue) and aversive (red) state (A). Under optimistic learning conditions (B), all states except
  for the harmful one become positive. Under pessimistic conditions (C), negative value spreads
  from the source to its antecedent states. The extent of the spread is worse for increased
  pessimism (D). (Parameters: $\gamma = 0.95$)
\end{figure}

@@

Pessimistic inference in anxiety is well-documented. In studies of self-reported beliefs,
individuals with anxiety rate
hypothetical future negative outcomes as being more likely to occur and worse to
experience than non-anxious counterparts \citep{ButlerMathews1983, ButlerMathews1987,
Foa1996, MacLeod1996, MacLeod1997, Luten1997, Stober1997, Borkovec1999, Maner2006,
Mitte2007, Miranda2007}. In the laboratory, exaggerated threat appraisal in anxiety
has traditionally been measured through fear conditioning. The results of these
studies demonstrate that anxious individuals exhibit greater fear towards threatening
cues, non-threatening cues, and extinguished threatening cues as compared to non-anxious
counterparts \citep{lissek2005, MinekaOehlberg2008, Duits2015}. Importantly, threatening
cues can become associated with secondary neutral cues, allowing negative value to
spread citep{wessa2007}. In sum, anxiety is associated with inflated estimates of
threat, both likelihood and value, and these estimates can spread from the source
to distantly associated cues.

@@

The same cannot be said for a the pessimistic agent. If an agent doubts its future
ability to act in according with the optimal policy, then it cannot rule out that,
in the future, it may inadvertently encounter the harmful state. This is more so
likely if the agent is already proximal to the harmful state. Under this pessimistic
belief, the possibility of future harm is no longer zero, as it is for the optimistic
agent, and thus negative value can propagate from the states antecedent to harm,
back to their respective antecedents and so on. The extent of this spread will depend
on the strength of the pessimistic belief (Figures 1c, 1d). With increasing pessimism,
harmful outcomes are assigned greater credence and thus exert greater and wider
influence on antecedent states.

@@

A second symptom central to all anxiety disorders is avoidance \citep{dsm5,
Krypotos2015, Arnaudova2017}, or actions taken to increase or maintain the temporal,
physical and/or mental distance between an individual and a threatening situation.
Although avoidance is a natural response to danger, excessive avoidance as is
observed in anxiety disorders is can be highly disruptive to everyday functioning
\cite{Salter2004} and may work to sustain anxious appraisals and behavior. For
example, an individual with social anxiety may avoid attending large social events
for fear of public embarrassment only to never learn that such worry is not justified.

Avoidance is often chosen even at the expense of potential reward (approch-avoidance
conflict), and is frequently a major impediment to recovery \citep{Arnaudova2017}.

In the laboratory, avoidance has been measured in a variety of ways. In animal
models of anxiety, anxious rodents are found to exhibit avoidance learning more
quickly and maintain avoidance for longer \citep{servatius2008}. In studies of
human avoidance, anxiety has been found to correlate with the maintenance of
increased distance from threat, such as virtual predators \citep{Bach2014, Bach2017,
Sheynin2014}.

% Arnaudova2017 makes reference to some spider approach studies that might be
% interesting to try and track down. Essential idea is that even though
% participants are totally safe, may not touch the glass. The task is the
% behavioral avoidance task. (e.g. Siegel & Weinberger, 2012, emotion).
% the nice part about this is that participants are objectively safe.

@@

% Need to go into more detail than this, need to actually make this a coherent
% explanation, but this can wait for now.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/02_cliff.png}}%
  }
  \caption{\textbf{Cliff Walking}}
  \par Cliff = -100, Cost = -1, gamma = 0.99
\end{figure}

In our second result, we demonstrate how a breakdown in control can lead to
sustained policies of avoidance. We do so in the cliff-walking environment
\citep{SuttonBarto1998, SuttonBarto2018, Gaskett2003}. Here an agent learns to navigate from
an initial starting position to an end state. The shortest distance is along a
cliff edge. If the agent steps off the cliff, it incurs a large penalty. Under
complete control (objective or perceived), agent in this environment is under no
threat despite the presence of the cliff's edge. As in the first simulation, this
is simply because, in a fully controllable world, an agent need not ever interact
with the aversive state. In fact, an agent can decide to walk straight up to the
aversive state and, insofar it does not choose to take one step further, will
never be in any actual danger.

This is not so when deterministic control cannot be assumed. If an agent believes
it is not in full control of its environment, then it cannot rule out the possibility
that it may inadvertently encounter the negative state at some point in the future,
especially if it veers to close to the state. This is true irrespective of whether
the agent believes it cannot guarantee its own future competency or the benevolence
of the environment. Importantly, this belief need not match the true environmental
statistics; belief is enough.

The long-run state values under different beliefs are plotted in Figure 2. As
expected, under optimistic beliefs (Figure 1b) the entire environment sans negative
state takes on a positive value (close to the positive state itself). As this belief
breaks down, we observe a different pattern in the long-run state values. When
future occupancy of the negative state becomes a non-zero probability, the undesired
action transition exerts influence on action-outcome calculations, back-propagating,
and thereby infecting antecedent states. As the agent's belief becomes increasingly
pessimistic, the more the worst possible action exerts influence and the more
negative, on average, the state-space grows. Most importantly, because these are
the long-run state values, the policy is the optimal policy; in other words, this
is the long-run policy an agent will take.

>>>>

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/01_field.png}}%
  }
  \par \textbf{Figure 1:} The open field is a deterministic gridworld with only one rewarding
  (blue) and aversive (red) state (A). For an optimistic agent ($w=1$), all states (other than
  the harmful state) take on positive value (B). For a pessimistic agent ($w=0.5$), negative value spreads
  from the source to antecedent states (C). With increasing pessimism ($w=0$), the extent of the spread
  grows worse (D). (Parameters: $\gamma = 0.95$)
\end{figure}

@@

Pessimistic inference in anxiety is well-documented. In studies of self-reported beliefs,
individuals with anxiety rate
hypothetical future negative outcomes as being more likely to occur and worse to
experience than non-anxious counterparts \citep{ButlerMathews1983, ButlerMathews1987,
Foa1996, MacLeod1996, MacLeod1997, Luten1997, Stober1997, Borkovec1999, Maner2006,
Mitte2007, Miranda2007}. In the laboratory, exaggerated threat appraisal in anxiety
has traditionally been measured through fear conditioning. The results of these
studies demonstrate that anxious individuals exhibit greater fear towards threatening
cues, non-threatening cues, and extinguished threatening cues as compared to non-anxious
counterparts \citep{lissek2005, MinekaOehlberg2008, Duits2015}. Importantly, threatening
cues can become associated with secondary neutral cues, allowing negative value to
spread \citep{wessa2007}. In sum, anxiety is associated with inflated estimates of
threat, both likelihood and value, and these estimates can spread from the source
to distantly associated cues.

@@

The same cannot be said for a pessimistic agent. If an agent doubts its future
ability to act in according with the optimal policy, then it cannot rule out that,
in the future, it may inadvertently encounter the harmful state. This is more so
likely if the agent is already proximal to the harmful state. Under this pessimistic
belief, the possibility of future harm is no longer zero, as it is for the optimistic
agent, and thus negative value can propagate from the states antecedent to harm,
back to their respective antecedents and so on. The extent of this spread will depend
on the strength of the pessimistic belief (Figures 1c, 1d). With increasing pessimism,
harmful outcomes are assigned greater credence and thus exert greater and wider
influence on antecedent states.

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/02_cliff.png}}%
  }
  \par \textbf{Figure 2:} Cliff-walking is a deterministic gridworld where an agent
  must navigate from one side (S) to the goal on the other side (G) (A). Every
  step costs the agent ($r=-1$) but stepping off the cliff results in great harm ($r=-100$).
  An optimistic agent learns to travel the shortest (loss-minimizing) path along the cliff (B).
  For a pessimistic agent ($w=0.8$), the potential for harm yields an optimal policy that
  maintains distance from the cliff's edge (C). With increasing pessimism ($w=0.6$), the agent
  learns an optimal path off the cliff so as to minimize unnecessary steps prior to its
  perceived inevitable harm (D). (Parameters: $\gamma = 1$)
\end{figure}

In addition to pessimistic inference, all anxiety disorders share pathological
avoidance as core symptom \citep{dsm5, Krypotos2015, Arnaudova2017}. Avoidance
behaviors seekto increase or maintain the temporal, physical and/or mental distance
between an individual and a threatening situation. Although avoidance is a natural response to danger, excessive avoidance as is observed in anxiety disorders can be highly disruptive to everyday
functioning \cite{Salter2004} and frequently function so as maintain anxiety
by preventing extinction \citep{Arnaudova2017}. For example, an individual with social anxiety may avoid
attending large social events for fear of public embarrassment only to never learn
that such worry is not justified.

Avoidance behaviors in anxiety have been studied using a variety of methods. Perhaps
the most classic paradigm is the behavioral avoidance test \citep{bandura1977}, wherein participants
complete a number of tasks involving a feared object or situation and the number of
tasks completed (or lack thereof) is a measure of avoidance. Computerized avoidance
tasks have become increasingly common, and increased distance from virtual threat
has been found to correlate with anxiety \citep{Bach2014, Bach2017, Sheynin2014}.

@@

In our second simulation, we demonstrate how pessimistic learning can yield
sustained avoidance. In the cliff-walking environment \citep{SuttonBarto1998,
SuttonBarto2018, Gaskett2003}, an agent must navigate from a start on side of the
map to the other (Figure 2a). The agent experiences a small penalty for every
step it takes. The shortest (least costly) path is along the cliff's edge; if the
agent steps off the cliff, however, it incurs a much larger penalty. As in the first
environment, we assume a deterministic transition structure.

Just as in the first environment, the long-run value estimates by an optimistic
agent do not reflect the potential harm of the cliff (Figure 2b). Because this agent (rightfully)
expects to act according to the optimal policy in the future (i.e. never step off
the cliff), the agent perceives no risk of future disaster and thus the large
penalties of stepping off the cliff do not pollute the agent's value estimates.
The optimistic agent learns the loss-minimizing policy and walks along the cliff's edge.

This is not the case for the pessimistic agent. As above, when the agent doubts
its future ability to act in a reward-maximizing (loss-minimizing) manor, it
cannot rule out the possibility of accidentally stepping off the cliff. Thus,
the potential for future harm is no longer zero and the large penalties for stepping
off the cliff may back-propagate to their antecedent states, and so on. The extent
of this spread is dependent on the strength of the pessimistic belief.
With increasing pessimism, the optimal policy for the agent is increasingly
distance from the cliff's edge (Figure 2c). Surprisingly, at high levels of pessimistic
belief, the optimal policy is to immediately step off the cliff (Figure 2d). The
reason for this is simple: with high levels of pessimism, the likelihood of disaster
grows large enough that any step is simply viewed as adding onto the pain.

Under pessimistic learning then, a perceived lack of self-efficacy can yield the
propagation of negative value from potential future harms. In such a scenario,
the optimal policy for an agent is to maintain a distance from threat far enough
so as to optimally balance the risk of catastrophe with foregone losses. Importantly,
these simulations demonstrate the long-run value estimates rather than any momentary
value estimates. In other words, the behavioral policy these agents demonstrate
will be maintained insofar beliefs about self-efficacy are maintained, irrespective
of any momentary noisiness in actual outcomes. Thus, pessimistic learning can
account for pathological avoidance in anxiety disorders.
